---
title: "M09-2-Application: Data import and export"
author: Allen Cosme
date: today
format: 
    html:
        theme: cosmo
        toc: true
        toc-location: right-body
        toc-depth: 4
        code-line-numbers: true
        code-link: true
        number-sections: true
        embed-resources: true
        df-print: paged
Executed: 
    freeze: true
    echo: fenced
    warning: false
---

# Preparation

```{r}
# install.packages("tidyverse")
# install.packages("readxl")
# install.packages("readxl")
library(tidyverse)
library(readxl)
library(readxl)
```

#  Dealing with csv file

## Importing csv file

### 

```{r}
item_csv <- read_csv2("/Users/akcosme/Downloads/items_ordered.csv",
                      trim_ws = TRUE,
                      show_col_types = FALSE)

glimpse(item_csv)
```

```{r}
item_csv <- item_csv %>%
  mutate(
    order_date = dmy(order_date),   # or mdy/order_date depending on format
    quantity   = as.integer(quantity),
  )

glimpse(item_csv)
```

## 

```{r}
library(readr)
library(dplyr)
library(lubridate)

item_csv <- read_csv2(
  "/Users/akcosme/Downloads/items_ordered.csv",
  trim_ws = TRUE,
  show_col_types = FALSE
) %>%
  mutate(
    order_date = dmy(order_date)
  )

most_expensive_item <- item_csv %>%
  filter(price == max(price, na.rm = TRUE)) %>%
  pull(item)

most_expensive_item
```
# dealing with an excel file

##
```{r}
library(readxl)
library(dplyr)
library(ggplot2)

item_excel <- read_excel("/Users/akcosme/Downloads/items_ordered.xlsx",
                         trim_ws = TRUE)
glimpse(item_excel)
```

##

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

item_excel <- read_excel("/Users/akcosme/Downloads/items_ordered.xlsx",
                         trim_ws = TRUE)

if (!dir.exists("images")) {
  dir.create("images")
}

p <- ggplot(item_excel, aes(x = item, y = quantity)) +
  geom_col() +
  labs(
    title = "Quantity of Items Ordered",
    x = "Item",
    y = "Quantity"
  )

ggsave("images/items_quantity.png", plot = p, width = 8, height = 6)
ggsave("images/items_quantity.pdf", plot = p, width = 8, height = 6)
ggsave("images/items_quantity.jpeg", plot = p, width = 8, height = 6)
```

# Importing from the clipboard and cleaning NA's or "--"

##

```{r}
library(dplyr)
library(janitor)
library(readr)
library(writexl)

dis_exec <- read.delim(pipe("pbpaste"), sep = "\t", header = TRUE)

dis_exec_clean <- dis_exec %>%
  select(where(~ !all(is.na(.)))) %>%   # remove empty columns
  mutate(across(where(is.character), ~ na_if(., "N/A"))) %>%
  mutate(across(where(is.character), ~ na_if(., "---"))) %>%
  clean_names()

if (!dir.exists("data")) {
  dir.create("data")
}

write_csv(dis_exec_clean, "data/dis_exec_clean.csv")
write_xlsx(dis_exec_clean, "data/dis_exec_clean.xlsx")
```

# web scraping

##

```{r}
library(xml2)
library(rvest)
library(stringr)
library(dplyr)
library(tibble)

page <- read_html("/Users/akcosme/Downloads/IMDb Top 250 movies.html")
html_text <- readLines("/Users/akcosme/Downloads/IMDb Top 250 movies.html", warn = FALSE) |>
  paste(collapse = " ")

top_movies <- str_match_all(html_text, '"titleText":\\{"text":"([^"]+)"\\}')[[1]][,2]
top_movies <- top_movies[top_movies != ""]
top_movies <- unique(top_movies)[1:250]
```

##

```{r}
year <- page %>%
  html_elements("your_selector_here") %>%
  html_text() %>%
  str_trim()

year <- year[str_detect(year, "\\d{4}")]
```

##

```{r}
year <- str_match_all(html_text, '"releaseYear":\\{"year":([0-9]{4})')[[1]][,2]
year <- year[year != ""]
year <- year[1:250]

length(year)
year
```

##

```{r}
rating <- str_match_all(html_text, '"aggregateRating":([0-9]\\.[0-9])')[[1]][,2]
rating <- rating[rating != ""]
rating <- rating[1:250]

length(rating)
rating
```

# Creating a data frame from web-scraped data

```{r}
library(dplyr)
library(tibble)
library(stringr)

# clean vectors first if needed
year <- str_extract(year, "\\d{4}")
rating <- str_extract(rating, "\\d+\\.?\\d*")

# check lengths
length(top_movies)
length(year)
length(rating)

# create data frame
top250_movies <- tibble(
  top_movies = top_movies,
  year = year,
  rating = as.numeric(rating)
) %>%
  arrange(desc(rating))

# print top and bottom 6 rows
head(top250_movies, 6)
tail(top250_movies, 6)
```

# Replacing values with NA

```{r}
# install.packages("naniar")
library(tibble)
library(naniar)
library(dplyr)

data <- tibble::tribble(
  ~name,            ~x,  ~y,              ~z,
  "N/A",             1,  "N/A",          -100,
  "N A",             3,  "NOt available",-99,
  "N / A",          NA,  "29",           -98,
  "Not Available", -99,  "25",          -101,
  "John Smith",    -98,  "28",            -1
)

data
```

##

```{r}
q6_1 <- data %>%
  replace_with_na_all(condition = ~ .x == "N/A")

q6_1
```

##

```{r}
q6_2_a <- data %>%
  replace_with_na(replace = list(
    name = c("N/A", "N A", "N / A"),
    y = c("N/A", "N A", "N / A")
  ))

q6_2_a
```

```{r}
q6_2_b <- data %>%
  replace_with_na_all(condition = ~ .x %in% c("N/A", "N A", "N / A"))

q6_2_b
```

```{r}
q6_2_c <- data %>%
  replace_with_na_all(condition = ~ str_detect(.x, "N\\s*/?\\s*A"))

q6_2_c
```

```{r}
q6_3 <- data %>%
  replace_with_na_all(condition = ~ str_detect(.x, "N\\s*/?\\s*A|Not Available"))

q6_3
```

```{r}
q6_2_d <- data %>%
  replace_with_na_at(
    .vars = c("name", "y"),
    condition = ~ .x %in% c("N/A", "N A", "N / A")
  )

q6_2_d
```

# Impact of GDP on student/teacher ratio

```{r}
# install.packages("WDI")
library(tidyverse)
library(WDI)
library(ggrepel)
library(scales)
```

##

```{r}
student_teacher <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-07/student_teacher_ratio.csv"
)

glimpse(student_teacher)
summary(student_teacher)

# 1) variables with missing values
colSums(is.na(student_teacher))

# 2) years covered and their distribution
student_teacher %>%
  count(year) %>%
  arrange(year)

# 3) all indicators and their distribution
student_teacher %>%
  count(indicator) %>%
  arrange(desc(n))

# filter data
s_t_ratio_2nd_edu_2016 <- student_teacher %>%
  filter(
    year == 2016,
    indicator == "Secondary Education",
    !is.na(student_ratio)
  )

dim(s_t_ratio_2nd_edu_2016)
glimpse(s_t_ratio_2nd_edu_2016)
```

```{r}
unique(student_teacher$indicator)
```

##

```{r}
WDIsearch("gdp per capita")
WDIsearch("population total")
```

##

```{r}
indicators_2016 <- WDI(
  country = "all",
  indicator = c(gdp_per_capita = "NY.GDP.PCAP.CD", population = "SP.POP.TOTL"),
  start = 2016,
  end = 2016,
  extra = TRUE
) %>%
  filter(!is.na(iso3c)) %>%
  select(iso3c, country, region, gdp_per_capita, population)
```

```{r}
# create iso3c if missing (use country_code when it's 3-letter; otherwise map from country names)
if (!"iso3c" %in% names(s_t_ratio_2nd_edu_2016)) {
  sample_codes <- head(na.omit(s_t_ratio_2nd_edu_2016$country_code), 50)
  if (length(sample_codes) > 0 && mean(nchar(sample_codes) == 3) > 0.5) {
    s_t_ratio_2nd_edu_2016 <- s_t_ratio_2nd_edu_2016 |> dplyr::mutate(iso3c = country_code)
  } else {
    if (!requireNamespace("countrycode", quietly = TRUE)) install.packages("countrycode")
    s_t_ratio_2nd_edu_2016 <- s_t_ratio_2nd_edu_2016 |> dplyr::mutate(iso3c = countrycode::countrycode(country, "country.name", "iso3c"))
  }
}

ratio_gdp_2016 <- s_t_ratio_2nd_edu_2016 |> dplyr::left_join(indicators_2016, by = "iso3c")

ratio_gdp_2016
```

```{r}
ggplot(ratio_gdp_2016, aes(x = student_ratio)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Student/Teacher Ratio",
    x = "Student/Teacher Ratio",
    y = "Count"
  )
```

```{r}
ggplot(ratio_gdp_2016, aes(x = student_ratio)) +
  geom_histogram(bins = 30) +
  scale_x_log10() +
  labs(
    title = "Distribution of Student/Teacher Ratio (Log X-Axis)",
    x = "Student/Teacher Ratio (log scale)",
    y = "Count"
  )
```

##

```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(scales)

plot_data <- ratio_gdp_2016 %>%
  mutate(country_label = coalesce(country.x, country.y)) %>%
  filter(
    !is.na(gdp_per_capita),
    !is.na(population),
    !is.na(region),
    !is.na(country_label),
    !is.na(student_ratio)
  )

ggplot(
  plot_data,
  aes(
    x = gdp_per_capita,
    y = student_ratio,
    size = population,
    color = region
  )
) +
  geom_point(alpha = 0.7) +
  geom_text_repel(
    aes(label = country_label),
    size = 3,
    max.overlaps = 100,
    show.legend = FALSE
  ) +
  geom_smooth(
    aes(x = gdp_per_capita, y = student_ratio),
    method = "lm",
    se = FALSE,
    color = "black",
    inherit.aes = FALSE
  ) +
  scale_x_log10(labels = label_dollar()) +
  scale_size_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Impact of GDP per Capita on Student/Teacher Ratio",
    subtitle = "Secondary Education, 2016",
    x = "GDP per Capita (current US$, log scale)",
    y = "Student/Teacher Ratio",
    color = "Region",
    size = "Population",
    caption = "Sources: TidyTuesday student-teacher ratio data and World Bank WDI"
  ) +
  theme_minimal()
```

# Labelled data

```{r}
# install.packages("labelled")
library(haven)
library(labelled)
library(dplyr)

demo <- read_sav("/Users/akcosme/Downloads/demo.sav")

look_for(demo)

glimpse(demo)
```

##

```{r}
library(janitor)

tabyl(as_factor(demo$employ)) |> adorn_pct_formatting()
tabyl(as_factor(demo$retire)) |> adorn_pct_formatting()
tabyl(as_factor(demo$empcat)) |> adorn_pct_formatting()
tabyl(as_factor(demo$jobsat)) |> adorn_pct_formatting()

summary(demo$jobsat)
```

##

```{r}
library(tidyr)
library(ggplot2)

tech_counts <- demo %>%
  select(wireless, multline, voice, pager, internet, callid, callwait) %>%
  mutate(across(everything(), as_factor)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "technology",
    values_to = "status"
  ) %>%
  filter(!is.na(status)) %>%
  count(technology, status)

tech_counts
```

```{r}
ggplot(tech_counts, aes(x = technology, y = n, fill = status)) +
  geom_col(position = "dodge") +
  labs(
    title = "Job-Related Technology Ownership",
    x = "Technology Product",
    y = "Count",
    fill = "Status"
  ) +
  theme_minimal()
```

# Impact of retirement on the use of the total job-related technology products

##

```{r}
library(dplyr)
library(haven)
library(labelled)

demo <- demo %>%
  mutate(
    total_job_tech =
      (as.numeric(as_factor(wireless)) == 2) +
      (as.numeric(as_factor(multline)) == 2) +
      (as.numeric(as_factor(voice)) == 2) +
      (as.numeric(as_factor(pager)) == 2) +
      (as.numeric(as_factor(internet)) == 2) +
      (as.numeric(as_factor(callid)) == 2) +
      (as.numeric(as_factor(callwait)) == 2)
  )

var_label(demo$total_job_tech) <- "Total number of job-related technology products used"

summary(demo$total_job_tech)
```

##

```{r}
library(janitor)

demo %>%
  mutate(
    retire = as_factor(retire),
    empcat = as_factor(empcat),
    jobsat = as_factor(jobsat)
  ) %>%
  group_by(retire, empcat, jobsat) %>%
  summarise(
    avg_total_job_tech = mean(total_job_tech, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

##

Retirement status, employment category, and job satisfaction appear to be related to the average number of job-related technology products used. Groups that are more actively connected to work tend to use more job-related technology, while retired or less work-involved groups tend to use fewer. The exact pattern should be interpreted from the averages shown in the tables.

# Job-related technology product ownership

```{r}
library(tidyr)
library(ggplot2)
library(scales)
library(forcats)

tech_prop <- demo %>%
  transmute(
    wireless = as_factor(wireless) == "Yes",
    multline = as_factor(multline) == "Yes",
    voice = as_factor(voice) == "Yes",
    pager = as_factor(pager) == "Yes",
    internet = as_factor(internet) == "Yes",
    callid = as_factor(callid) == "Yes",
    callwait = as_factor(callwait) == "Yes"
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "technology",
    values_to = "owns"
  ) %>%
  group_by(technology) %>%
  summarise(
    proportion = mean(owns, na.rm = TRUE),
    n = sum(!is.na(owns)),
    se = sd(as.numeric(owns), na.rm = TRUE) / sqrt(n),
    lower = proportion - 1.96 * se,
    upper = proportion + 1.96 * se,
    .groups = "drop"
  ) %>%
  mutate(
    technology = fct_reorder(technology, proportion, .desc = TRUE),
    lower = pmax(lower, 0),
    upper = pmin(upper, 1)
  )

ggplot(tech_prop, aes(x = technology, y = proportion)) +
  geom_col() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  geom_text(aes(label = percent(proportion, accuracy = 0.1)), vjust = -0.4) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1.05)) +
  labs(
    title = "Proportion Owning Each Job-Related Technology Product",
    subtitle = "Bars show sample proportions with 95% confidence intervals",
    x = "Technology Product",
    y = "Proportion Owning Product",
    caption = "Error bars: proportion +/- 1.96 x standard error"
  ) +
  theme_minimal()
```

##
The bar chart shows the proportion of respondents who own each job-related technology product. The products are ordered from highest to lowest proportion, and the error bars represent 95% confidence intervals around each estimated proportion.
