---
title: "M09-1-Principle: Data import and export"
author: Allen Cosme
date: today
format: 
    html:
        theme: cosmo
        toc: true
        toc-location: right-body
        toc-depth: 4
        code-line-numbers: true
        code-link: true
        number-sections: true
        embed-resources: true
        df-print: paged
Executed: 
    freeze: true
    echo: fenced
    warning: false
---

::: callout-note
**note** the lecture notes will be available on [Professor Jungs quarto page](https://jmjung.quarto.pub/m09-1-principle-data-import-and-export/)
:::

# Overview

## Expected Learning Outcomes
- Explain how to create a Github repository and collaborate with others on the same R projects.
- Effectively load and look through built-in datasets in R.
- Import various type of data (csv, xlsx, and SPSS) to RStudio.
- Scrape data from the web using SelectorGadget, rvest, and inspect function of web browers.
- Import multiple data sets and work with them.
- Work with labelled data in R.
- Export/save output data to local pc and push to Github.

## Textbook chapters covered

- ch. 7: Data Import
- ch. 20: Speadsheets
- ch. 24: Web scraping
- ch 18: Missing values

# Installing and loading up R Packages
```{r}
#| label: package loading

library(tidyverse)
library(sjPlot)
```

# Reading Built in datasets
```{r}
# look at data sets in the loaded packages
data()

# geting information about the lakers dataset
head(lakers)
lakers$player
lakers |>
    as_tibble()
```

# importing data

## `csv` files
```{r}
#read .csv files
stu_per_base <- read.csv("/Users/akcosme/Downloads/StudentsPerformance.csv")

class(stu_per_base)
head(stu_per_base)
```

## using `base r`
```{r}
#?read.csv
stu_per_base <- read.csv("/Users/akcosme/Downloads/StudentsPerformance.csv") # Not as good as read_csv

class(stu_per_base)
head(stu_per_base)
```

## using `readr`

```{r}
stu_per_readr <- read_csv("/Users/akcosme/Downloads/StudentsPerformance.csv") # Use read_csv() all the time from now on over read.csv()

spec(stu_per_readr)
class(stu_per_readr)
head(stu_per_readr)

skimr::skim(stu_per_readr) #  better than summary() function

stu_per_readr$`math score` # RStudio's auto prompt
stu_per_readr$"math score" # same outcome
stu_per_readr$'math score'
```

# Importing `Excel` data with `readxl` package

```{r readxl}
library(readxl)

sp_excel <- read_excel("/Users/akcosme/Downloads/StudentsPerformance.xlsx") 

#?read_excel

sp_excel
skimr::skim(sp_excel)
```

## Importing data from the web

```{r}
read_csv("https://raw.githubusercontent.com/jaejungca/R-data-import-export/refs/heads/main/data/StudentsPerformance.csv")
```